---
title: "QTL Mapping of Chicken Comb Mass"
subtitle: "An F8 advanced intercross practical"
author: "Your name"
date: today
format: html
editor: visual
---

## 1. Introduction {.unnumbered}

This practical investigates genetic regions associated with comb mass in an F8 advanced intercross between Red Junglefowl and White Leghorn chickens.

The objectives are to inspect the dataset, assess marker order and the genetic map, identify QTL, estimate their positions and effects, and compare results with and without adjustment for sex and body weight.

The supplied marker data cover chromosomes 1 and 3. Significance thresholds will therefore apply to the scanned chromosomes.

## 2. Load and inspect data

```{r}
#| label: locate-input-data

data_path <- "data/raw/QTL_group_work.csv"
stopifnot(
  file.exists(data_path)
)

tools::md5sum(data_path)
```

```{r}
#| label: inspect-dataset-layout

raw_csv <- read.csv(
  data_path,
  header = FALSE,
  colClasses = "character",
  na.strings = NULL,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dim(raw_csv)

raw_csv[1:10, 1:10]
```

```{r}
dplyr::glimpse(raw_csv)
```

#### Observations

The dataset contains 575 rows and 44 columns. The first three rows contain column names, chromosome assignments, and supplied map positions, leaving 572 individual records. The first four columns contain individual ID, body weight at 212 days, sex, and comb mass; the remaining 40 columns contain marker genotypes. The displayed records include genotype codes A, H, and B. All entries were initially imported as text to preserve the original coding. The supplied position values require further assessment before being interpreted as genetic distances.

### Separate individual records and marker information

```{r}
#| label: separate-input-records

column_names <- as.character(
  unlist(raw_csv[1, ], use.names = FALSE)
)

stopifnot(
  !anyDuplicated(column_names),
  identical(column_names[1:4], c("id", "w212", "sex", "comb_g"))
)

individual_records <- raw_csv[-c(1, 2, 3), , drop = FALSE]
names(individual_records) <- column_names

marker_columns <- 5:ncol(raw_csv)

marker_metadata <- data.frame(
  marker = column_names[marker_columns],
  chromosome = as.character(
    unlist(raw_csv[2, marker_columns], use.names = FALSE)
  ),
  supplied_position = as.numeric(
    unlist(raw_csv[3, marker_columns], use.names = FALSE)
  )
)

cat("Individual records:")
nrow(individual_records)
cat("\nMarkermetadata")
nrow(marker_metadata)
cat("\nchromosome")
table(marker_metadata$chromosome)

stopifnot(
  !anyDuplicated(individual_records$id),
  !any(trimws(individual_records$id) %in% c("", "-", "NA"))
)
```

### Inspect genotype codes and missing phenotypes

```{r}
#| label: inspect-codes-and-missingness

genotype_codes <- unlist(
  individual_records[, marker_columns, drop = FALSE],
  use.names = FALSE
)

table(genotype_codes, useNA = "ifany")

phenotype_names <- c("w212", "sex", "comb_g")

missing_counts <- vapply(
  individual_records[, phenotype_names, drop = FALSE],
  function(values) sum(trimws(values) %in% c("", "-", "NA")),
  integer(1)
)

data.frame(
  Phenotype = phenotype_names,
  Missing = unname(missing_counts),
  Observed = nrow(individual_records) - unname(missing_counts)
)

table(individual_records$sex, useNA = "ifany")
```

### Observations

The dataset contains 572 individuals and 40 markers, with 9 markers on chromosome 1 and 31 on chromosome 3. Genotype entries comprise 8,045 A calls, 3,723 B calls, 10,031 H calls, 239 C calls, and 842 entries coded as missing (-). Missing entries represent 3.68% of all marker–individual combinations. The meaning of code C needs to be established.

Comb mass is available for 447 individuals, matching the number of birds with comb measurements described in the practical. Body weight is available for 471 individuals, and sex for 571. Sex is recorded as code 1 for 272 individuals and code 2 for 299 individuals, with one missing value; the labels male and female have not yet been assigned.

The numbers of available measurements do not establish how many individuals have all three variables recorded. This overlap needs to be checked before comparing scans with different covariates, so that changes in QTL results can be distinguished from changes in the analyzed sample.

### Check availability for covariate comparisons

```{r}
#| label: inspect-complete-phenotypes

phenotype_available <- sapply(
  individual_records[, c("comb_g", "sex", "w212")],
  function(values) {
    !is.na(values) & !(trimws(values) %in% c("", "-", "NA"))
  }
)

availability_summary <- data.frame(
  Required_variables = c(
    "Comb mass",
    "Comb mass and sex",
    "Comb mass and body weight",
    "Comb mass, sex, and body weight"
  ),
  Individuals = c(
    sum(phenotype_available[, "comb_g"]),
    sum(
      phenotype_available[, "comb_g"] &
        phenotype_available[, "sex"]
    ),
    sum(
      phenotype_available[, "comb_g"] &
        phenotype_available[, "w212"]
    ),
    sum(rowSums(phenotype_available) == 3)
  )
)

print(availability_summary, row.names = FALSE)
```

### Observations

All 447 individuals with recorded comb mass also have recorded sex. Of these, 440 have body-weight measurements, leaving seven individuals with comb mass and sex but no body weight. The primary comparison of unadjusted, sex-adjusted, body-weight-adjusted, and jointly adjusted scans will therefore use the same 440 individuals. This keeps sample composition constant, allowing differences between scans to be attributed to covariate adjustment rather than the inclusion or exclusion of birds.
