---
title: "QTL Mapping of Chicken Comb Mass"
subtitle: "An F8 advanced intercross practical"
author: "Your name"
date: today
format: html
editor: visual
---

## 0. Load libraries

```{r}
library("qtl")
```

## 1. Introduction {.unnumbered}

This practical investigates genetic regions associated with comb mass in an F8 advanced intercross between Red Junglefowl and White Leghorn chickens.

The objectives are to inspect the dataset, assess marker order and the genetic map, identify QTL, estimate their positions and effects, and compare results with and without adjustment for sex and body weight.

The supplied marker data cover chromosomes 1 and 3. Significance thresholds will therefore apply to the scanned chromosomes.

## 2. Load and inspect data

```{r}
#| label: locate-input-data

data_path <- "data/raw/QTL_group_work.csv"
stopifnot(
  file.exists(data_path)
)

tools::md5sum(data_path)
```

```{r}
#| label: inspect-dataset-layout

raw_csv <- read.csv(
  data_path,
  header = FALSE,
  colClasses = "character",
  na.strings = NULL,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dim(raw_csv)

raw_csv[1:10, 1:10]
```

```{r}
dplyr::glimpse(raw_csv)
```

#### Observations

The dataset contains 575 rows and 44 columns. The first three rows contain column names, chromosome assignments, and supplied map positions, leaving 572 individual records. The first four columns contain individual ID, body weight at 212 days, sex, and comb mass; the remaining 40 columns contain marker genotypes. The displayed records include genotype codes A, H, and B. All entries were initially imported as text to preserve the original coding. The supplied position values require further assessment before being interpreted as genetic distances.

### Separate individual records and marker information

```{r}
#| label: separate-input-records

column_names <- as.character(
  unlist(raw_csv[1, ], use.names = FALSE)
)

stopifnot(
  !anyDuplicated(column_names),
  identical(column_names[1:4], c("id", "w212", "sex", "comb_g"))
)

individual_records <- raw_csv[-c(1, 2, 3), , drop = FALSE]
names(individual_records) <- column_names

marker_columns <- 5:ncol(raw_csv)

marker_metadata <- data.frame(
  marker = column_names[marker_columns],
  chromosome = as.character(
    unlist(raw_csv[2, marker_columns], use.names = FALSE)
  ),
  supplied_position = as.numeric(
    unlist(raw_csv[3, marker_columns], use.names = FALSE)
  )
)

cat("Individual records:")
nrow(individual_records)
cat("\nMarkermetadata")
nrow(marker_metadata)
cat("\nchromosome")
table(marker_metadata$chromosome)

stopifnot(
  !anyDuplicated(individual_records$id),
  !any(trimws(individual_records$id) %in% c("", "-", "NA"))
)
```

### Inspect genotype codes and missing phenotypes

```{r}
#| label: inspect-codes-and-missingness

genotype_codes <- unlist(
  individual_records[, marker_columns, drop = FALSE],
  use.names = FALSE
)

table(genotype_codes, useNA = "ifany")

phenotype_names <- c("w212", "sex", "comb_g")

missing_counts <- vapply(
  individual_records[, phenotype_names, drop = FALSE],
  function(values) sum(trimws(values) %in% c("", "-", "NA")),
  integer(1)
)

data.frame(
  Phenotype = phenotype_names,
  Missing = unname(missing_counts),
  Observed = nrow(individual_records) - unname(missing_counts)
)

table(individual_records$sex, useNA = "ifany")
```

### Observations

The dataset contains 572 individuals and 40 markers, with 9 markers on chromosome 1 and 31 on chromosome 3. Genotype entries comprise 8,045 A calls, 3,723 B calls, 10,031 H calls, 239 C calls, and 842 entries coded as missing (-). Missing entries represent 3.68% of all marker–individual combinations. The meaning of code C needs to be established.

Comb mass is available for 447 individuals, matching the number of birds with comb measurements described in the practical. Body weight is available for 471 individuals, and sex for 571. Sex is recorded as code 1 for 272 individuals and code 2 for 299 individuals, with one missing value; the labels male and female have not yet been assigned.

The numbers of available measurements do not establish how many individuals have all three variables recorded. This overlap needs to be checked before comparing scans with different covariates, so that changes in QTL results can be distinguished from changes in the analyzed sample.

### Check availability for covariate comparisons

```{r}
#| label: inspect-complete-phenotypes

phenotype_available <- sapply(
  individual_records[, c("comb_g", "sex", "w212")],
  function(values) {
    !is.na(values) & !(trimws(values) %in% c("", "-", "NA"))
  }
)

availability_summary <- data.frame(
  Required_variables = c(
    "Comb mass",
    "Comb mass and sex",
    "Comb mass and body weight",
    "Comb mass, sex, and body weight"
  ),
  Individuals = c(
    sum(phenotype_available[, "comb_g"]),
    sum(
      phenotype_available[, "comb_g"] &
        phenotype_available[, "sex"]
    ),
    sum(
      phenotype_available[, "comb_g"] &
        phenotype_available[, "w212"]
    ),
    sum(rowSums(phenotype_available) == 3)
  )
)

print(availability_summary, row.names = FALSE)
```

### Observations

All 447 individuals with recorded comb mass also have recorded sex. Of these, 440 have body-weight measurements, leaving seven individuals with comb mass and sex but no body weight. The primary comparison of unadjusted, sex-adjusted, body-weight-adjusted, and jointly adjusted scans will therefore use the same 440 individuals. This keeps sample composition constant, allowing differences between scans to be attributed to covariate adjustment rather than the inclusion or exclusion of birds.

## 3. Cross design, import, and validation

The study population is an F8 advanced intercross between Red Junglefowl and White Leghorn chickens. For this R/qtl practical, the data are represented as an intercross (`f2`).

This is an analytical approximation, not a reclassification of the population as biologically F2. The subsequent map assessment will consider accumulated recombination in the advanced intercross. The standard R/qtl workflow does not account for family relationships.

The import uses the standard R/qtl genotype convention: A = AA, H = AB, B = BB, D = AA or AB, and C = AB or BB.

### Import the cross

All 572 individuals are retained for marker and map assessment.

```{r}
#| label: import-comb-cross

comb_cross <- read.cross(
  format = "csv",
  dir = dirname(data_path),
  file = basename(data_path),
  genotypes = c("A", "H", "B", "D", "C"),
  alleles = c("A", "B"),
  na.strings = c("-", "NA"),
  crosstype = "f2",
  estimate.map = FALSE,
  convertXdata = FALSE
)
```

### Summarize the imported cross

```{r}
#| label: summarize-comb-cross

summary(comb_cross)

cat("\nCross object classes:\n")
class(comb_cross)

cat("\nNumber of individuals:\n")
nind(comb_cross)

cat("\nNumber of markers:\n")
totmar(comb_cross)

cat("\nNumber of markers per chromosome:\n")
nmar(comb_cross)

cat("\nPhenotype column names:\n")
names(comb_cross$pheno)
```

### Validate dimensions and individual order

```{r}
#| label: validate-cross-structure

stopifnot(
  inherits(comb_cross, "f2"),
  nind(comb_cross) == nrow(individual_records),
  totmar(comb_cross) == nrow(marker_metadata),
  setequal(names(comb_cross$geno), c("1", "3")),
  identical(
    as.character(comb_cross$pheno$id),
    as.character(individual_records$id)
  )
)

cat("Cross dimensions and individual order are consistent.\n")
```

### Validate genotype conversion

```{r}
#| label: validate-genotype-conversion

imported_genotypes <- unlist(
  lapply(comb_cross$geno, function(chromosome) {
    as.vector(chromosome$data)
  }),
  use.names = FALSE
)

cat("Imported genotype counts:\n")
table(
  factor(imported_genotypes, levels = 1:5),
  useNA = "ifany"
)

expected_counts <- c(
  sum(genotype_codes == "A"),
  sum(genotype_codes == "H"),
  sum(genotype_codes == "B"),
  sum(genotype_codes == "D"),
  sum(genotype_codes == "C")
)

observed_counts <- tabulate(
  imported_genotypes[!is.na(imported_genotypes)],
  nbins = 5
)

stopifnot(
  all(observed_counts == expected_counts),
  sum(is.na(imported_genotypes)) ==
    sum(genotype_codes %in% c("-", "NA", ""))
)

cat("\nGenotype conversion preserves the input counts.\n")
```

### Check imported phenotype missingness

```{r}
#| label: validate-imported-phenotypes

cat("Missing phenotype values after import:\n")
colSums(
  is.na(comb_cross$pheno[, c("comb_g", "sex", "w212")])
)

cat("\nImported sex codes:\n")
table(comb_cross$pheno$sex, useNA = "ifany")
```

#### **Observations and interpretation**

The imported cross contains 572 individuals and 40 markers across chromosomes 1 and 3, with 9 and 31 markers respectively. Validation confirmed that the dimensions and individual order match the original dataset.

Genotype conversion preserved 8,045 AA calls, 10,031 AB calls, 3,723 BB calls, 239 ambiguous AB-or-BB calls, and 842 missing entries. Under the adopted standard R/qtl coding convention, the original C entries were retained as ambiguous genotypes rather than discarded. Overall, 96.3% of marker–individual combinations have recorded genotype information.

Phenotype missingness was also preserved: 125 individuals lack comb mass, 101 lack body weight, and one lacks sex. Sex remains coded as 1 or 2 without assigning biological labels. Although R/qtl labels the object as an F2 intercross, this is an approximate analytical representation of the F8 population and does not account for family relationships.

## 3. Missing data and phenotype distributions

### Examine genotype missingness

Display genotype availability across individuals and markers. Black cells indicate missing calls; gray cells indicate partially informative genotype calls.

```{r}
#| label: fig-comb-missing-genotypes
#| fig-cap: "Genotype availability across chromosomes 1 and 3."
#| fig-height: 7

par(cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)

plotMissing(
  comb_cross,
  main = "Genotype availability in the comb dataset"
)
```

Quantify missing calls for each marker to identify the markers with the lowest data coverage.

```{r}
#| label: marker-missingness-summary

marker_missingness <- do.call(
  rbind,
  lapply(names(comb_cross$geno), function(chr) {
    genotype_matrix <- comb_cross$geno[[chr]]$data
    
    data.frame(
      chromosome = chr,
      marker = colnames(genotype_matrix),
      missing = colSums(is.na(genotype_matrix)),
      missing_percent = round(
        100 * colMeans(is.na(genotype_matrix)),
        2
      )
    )
  })
)

marker_missingness <- marker_missingness[
  order(-marker_missingness$missing),
]

# Calculate totals before adding the summary row.
total_missing <- sum(marker_missingness$missing)
total_entries <- nind(comb_cross) * nrow(marker_missingness)

marker_missingness <- rbind(
  marker_missingness,
  data.frame(
    chromosome = "All",
    marker = "Total",
    missing = total_missing,
    missing_percent = round(
      100 * total_missing / total_entries,
      2
    )
  )
)

rownames(marker_missingness) <- NULL

print(marker_missingness, row.names = FALSE)
```

Ambiguous AB-or-BB calls contain some genotype information and are not counted as fully missing in this table.

#### **Observations**

Overall, 842 of the 22,880 genotype entries are missing (3.68%). Missingness is concentrated at particular markers, especially on chromosome 3. Marker 3_23988469 has the highest missingness, with 259 of 572 individuals missing (45.28%), followed by Gg_rs15337358 (138; 24.13%), chr3_16000000 (84; 14.69%), and 3_16300000 (61; 10.66%). Together, these four markers account for 542 of the 842 missing entries (64.37%). On chromosome 1, 1_134165693 has the highest missingness (49; 8.57%). The remaining 35 markers each have no more than 5.07% missing genotypes.

#### **Interpretation**

The vertical bands in the plot indicate that genotype availability varies substantially between markers. Markers with high missingness provide less direct genotype information and warrant attention when examining marker order and estimating the genetic map. Because missingness alone does not establish a genotyping error or explain why the data are missing, these markers will not be removed solely on this evidence.
