---
title: "QTL Mapping of Chicken Comb Mass"
subtitle: "An F8 advanced intercross practical"
author: "Your name"
date: today
format: html
editor: visual
---

## 0. Load libraries

```{r}
library("qtl")
```

## 1. Introduction {.unnumbered}

This practical investigates genetic regions associated with comb mass in an F8 advanced intercross between Red Junglefowl and White Leghorn chickens.

The objectives are to inspect the dataset, assess marker order and the genetic map, identify QTL, estimate their positions and effects, and compare results with and without adjustment for sex and body weight.

The supplied marker data cover chromosomes 1 and 3. Significance thresholds will therefore apply to the scanned chromosomes.

## 2. Load and inspect data

```{r}
#| label: locate-input-data

data_path <- "data/raw/QTL_group_work.csv"
stopifnot(
  file.exists(data_path)
)

tools::md5sum(data_path)
```

```{r}
#| label: inspect-dataset-layout

raw_csv <- read.csv(
  data_path,
  header = FALSE,
  colClasses = "character",
  na.strings = NULL,
  check.names = FALSE,
  stringsAsFactors = FALSE
)

dim(raw_csv)

raw_csv[1:10, 1:10]
```

```{r}
dplyr::glimpse(raw_csv)
```

#### Observations

The dataset contains 575 rows and 44 columns. The first three rows contain column names, chromosome assignments, and supplied map positions, leaving 572 individual records. The first four columns contain individual ID, body weight at 212 days, sex, and comb mass; the remaining 40 columns contain marker genotypes. The displayed records include genotype codes A, H, and B. All entries were initially imported as text to preserve the original coding. The supplied position values require further assessment before being interpreted as genetic distances.

### Separate individual records and marker information

```{r}
#| label: separate-input-records

column_names <- as.character(
  unlist(raw_csv[1, ], use.names = FALSE)
)

stopifnot(
  !anyDuplicated(column_names),
  identical(column_names[1:4], c("id", "w212", "sex", "comb_g"))
)

individual_records <- raw_csv[-c(1, 2, 3), , drop = FALSE]
names(individual_records) <- column_names

marker_columns <- 5:ncol(raw_csv)

marker_metadata <- data.frame(
  marker = column_names[marker_columns],
  chromosome = as.character(
    unlist(raw_csv[2, marker_columns], use.names = FALSE)
  ),
  supplied_position = as.numeric(
    unlist(raw_csv[3, marker_columns], use.names = FALSE)
  )
)

cat("Individual records:")
nrow(individual_records)
cat("\nMarkermetadata")
nrow(marker_metadata)
cat("\nchromosome")
table(marker_metadata$chromosome)

stopifnot(
  !anyDuplicated(individual_records$id),
  !any(trimws(individual_records$id) %in% c("", "-", "NA"))
)
```

### Inspect genotype codes and missing phenotypes

```{r}
#| label: inspect-codes-and-missingness

genotype_codes <- unlist(
  individual_records[, marker_columns, drop = FALSE],
  use.names = FALSE
)

table(genotype_codes, useNA = "ifany")

phenotype_names <- c("w212", "sex", "comb_g")

missing_counts <- vapply(
  individual_records[, phenotype_names, drop = FALSE],
  function(values) sum(trimws(values) %in% c("", "-", "NA")),
  integer(1)
)

data.frame(
  Phenotype = phenotype_names,
  Missing = unname(missing_counts),
  Observed = nrow(individual_records) - unname(missing_counts)
)

table(individual_records$sex, useNA = "ifany")
```

### Observations

The dataset contains 572 individuals and 40 markers, with 9 markers on chromosome 1 and 31 on chromosome 3. Genotype entries comprise 8,045 A calls, 3,723 B calls, 10,031 H calls, 239 C calls, and 842 entries coded as missing (-). Missing entries represent 3.68% of all marker–individual combinations. The meaning of code C needs to be established.

Comb mass is available for 447 individuals, matching the number of birds with comb measurements described in the practical. Body weight is available for 471 individuals, and sex for 571. Sex is recorded as code 1 for 272 individuals and code 2 for 299 individuals, with one missing value; the labels male and female have not yet been assigned.

The numbers of available measurements do not establish how many individuals have all three variables recorded. This overlap needs to be checked before comparing scans with different covariates, so that changes in QTL results can be distinguished from changes in the analyzed sample.

### Check availability for covariate comparisons

```{r}
#| label: inspect-complete-phenotypes

phenotype_available <- sapply(
  individual_records[, c("comb_g", "sex", "w212")],
  function(values) {
    !is.na(values) & !(trimws(values) %in% c("", "-", "NA"))
  }
)

availability_summary <- data.frame(
  Required_variables = c(
    "Comb mass",
    "Comb mass and sex",
    "Comb mass and body weight",
    "Comb mass, sex, and body weight"
  ),
  Individuals = c(
    sum(phenotype_available[, "comb_g"]),
    sum(
      phenotype_available[, "comb_g"] &
        phenotype_available[, "sex"]
    ),
    sum(
      phenotype_available[, "comb_g"] &
        phenotype_available[, "w212"]
    ),
    sum(rowSums(phenotype_available) == 3)
  )
)

print(availability_summary, row.names = FALSE)
```

### Observations

All 447 individuals with recorded comb mass also have recorded sex. Of these, 440 have body-weight measurements, leaving seven individuals with comb mass and sex but no body weight. The primary comparison of unadjusted, sex-adjusted, body-weight-adjusted, and jointly adjusted scans will therefore use the same 440 individuals. This keeps sample composition constant, allowing differences between scans to be attributed to covariate adjustment rather than the inclusion or exclusion of birds.

## 3. Cross design, import, and validation

The study population is an F8 advanced intercross between Red Junglefowl and White Leghorn chickens. For this R/qtl practical, the data are represented as an intercross (`f2`).

This is an analytical approximation, not a reclassification of the population as biologically F2. The subsequent map assessment will consider accumulated recombination in the advanced intercross. The standard R/qtl workflow does not account for family relationships.

The import uses the standard R/qtl genotype convention: A = AA, H = AB, B = BB, D = AA or AB, and C = AB or BB.

### Import the cross

All 572 individuals are retained for marker and map assessment.

```{r}
#| label: import-comb-cross

comb_cross <- read.cross(
  format = "csv",
  dir = dirname(data_path),
  file = basename(data_path),
  genotypes = c("A", "H", "B", "D", "C"),
  alleles = c("A", "B"),
  na.strings = c("-", "NA"),
  crosstype = "f2",
  estimate.map = FALSE,
  convertXdata = FALSE
)
```

### Summarize the imported cross

```{r}
#| label: summarize-comb-cross

summary(comb_cross)

cat("\nCross object classes:\n")
class(comb_cross)

cat("\nNumber of individuals:\n")
nind(comb_cross)

cat("\nNumber of markers:\n")
totmar(comb_cross)

cat("\nNumber of markers per chromosome:\n")
nmar(comb_cross)

cat("\nPhenotype column names:\n")
names(comb_cross$pheno)
```

### Validate dimensions and individual order

```{r}
#| label: validate-cross-structure

stopifnot(
  inherits(comb_cross, "f2"),
  nind(comb_cross) == nrow(individual_records),
  totmar(comb_cross) == nrow(marker_metadata),
  setequal(names(comb_cross$geno), c("1", "3")),
  identical(
    as.character(comb_cross$pheno$id),
    as.character(individual_records$id)
  )
)

cat("Cross dimensions and individual order are consistent.\n")
```

### Validate genotype conversion

```{r}
#| label: validate-genotype-conversion

imported_genotypes <- unlist(
  lapply(comb_cross$geno, function(chromosome) {
    as.vector(chromosome$data)
  }),
  use.names = FALSE
)

cat("Imported genotype counts:\n")
table(
  factor(imported_genotypes, levels = 1:5),
  useNA = "ifany"
)

expected_counts <- c(
  sum(genotype_codes == "A"),
  sum(genotype_codes == "H"),
  sum(genotype_codes == "B"),
  sum(genotype_codes == "D"),
  sum(genotype_codes == "C")
)

observed_counts <- tabulate(
  imported_genotypes[!is.na(imported_genotypes)],
  nbins = 5
)

stopifnot(
  all(observed_counts == expected_counts),
  sum(is.na(imported_genotypes)) ==
    sum(genotype_codes %in% c("-", "NA", ""))
)

cat("\nGenotype conversion preserves the input counts.\n")
```

### Check imported phenotype missingness

```{r}
#| label: validate-imported-phenotypes

cat("Missing phenotype values after import:\n")
colSums(
  is.na(comb_cross$pheno[, c("comb_g", "sex", "w212")])
)

cat("\nImported sex codes:\n")
table(comb_cross$pheno$sex, useNA = "ifany")
```

#### **Observations and interpretation**

The imported cross contains 572 individuals and 40 markers across chromosomes 1 and 3, with 9 and 31 markers respectively. Validation confirmed that the dimensions and individual order match the original dataset.

Genotype conversion preserved 8,045 AA calls, 10,031 AB calls, 3,723 BB calls, 239 ambiguous AB-or-BB calls, and 842 missing entries. Under the adopted standard R/qtl coding convention, the original C entries were retained as ambiguous genotypes rather than discarded. Overall, 96.3% of marker–individual combinations have recorded genotype information.

Phenotype missingness was also preserved: 125 individuals lack comb mass, 101 lack body weight, and one lacks sex. Sex remains coded as 1 or 2 without assigning biological labels. Although R/qtl labels the object as an F2 intercross, this is an approximate analytical representation of the F8 population and does not account for family relationships.

## 4. Missing data and phenotype distributions

### Examine genotype missingness

Display genotype availability across individuals and markers. Black cells indicate missing calls; gray cells indicate partially informative genotype calls.

```{r}
#| label: fig-comb-missing-genotypes
#| fig-cap: "Genotype availability across chromosomes 1 and 3."
#| fig-height: 7

par(cex.main = 1.3, cex.lab = 1.2, cex.axis = 1.1)

plotMissing(
  comb_cross,
  main = "Genotype availability in the comb dataset"
)
```

Quantify missing calls for each marker to identify the markers with the lowest data coverage.

```{r}
#| label: marker-missingness-summary

marker_missingness <- do.call(
  rbind,
  lapply(names(comb_cross$geno), function(chr) {
    genotype_matrix <- comb_cross$geno[[chr]]$data
    
    data.frame(
      chromosome = chr,
      marker = colnames(genotype_matrix),
      missing = colSums(is.na(genotype_matrix)),
      missing_percent = round(
        100 * colMeans(is.na(genotype_matrix)),
        2
      )
    )
  })
)

marker_missingness <- marker_missingness[
  order(-marker_missingness$missing),
]

# Calculate totals before adding the summary row.
total_missing <- sum(marker_missingness$missing)
total_entries <- nind(comb_cross) * nrow(marker_missingness)

marker_missingness <- rbind(
  marker_missingness,
  data.frame(
    chromosome = "All",
    marker = "Total",
    missing = total_missing,
    missing_percent = round(
      100 * total_missing / total_entries,
      2
    )
  )
)

rownames(marker_missingness) <- NULL

print(marker_missingness, row.names = FALSE)
```

Ambiguous AB-or-BB calls contain some genotype information and are not counted as fully missing in this table.

#### **Observations**

Overall, 842 of the 22,880 genotype entries are missing (3.68%). Missingness is concentrated at particular markers, especially on chromosome 3. Marker 3_23988469 has the highest missingness, with 259 of 572 individuals missing (45.28%), followed by Gg_rs15337358 (138; 24.13%), chr3_16000000 (84; 14.69%), and 3_16300000 (61; 10.66%). Together, these four markers account for 542 of the 842 missing entries (64.37%). On chromosome 1, 1_134165693 has the highest missingness (49; 8.57%). The remaining 35 markers each have no more than 5.07% missing genotypes.

#### **Interpretation**

The vertical bands in the plot indicate that genotype availability varies substantially between markers. Markers with high missingness provide less direct genotype information and warrant attention when examining marker order and estimating the genetic map. Because missingness alone does not establish a genotyping error or explain why the data are missing, these markers will not be removed solely on this evidence.

### Extract and summarize phenotypes

```{r}
#| label: extract-comb-phenotypes

phenotype_data <- data.frame(
  id = comb_cross$pheno$id,
  comb_mass = pull.pheno(comb_cross, pheno.col = "comb_g"),
  body_weight = pull.pheno(comb_cross, pheno.col = "w212"),
  sex_code = pull.pheno(comb_cross, pheno.col = "sex")
)

phenotype_data$sex_group <- factor(
  phenotype_data$sex_code,
  levels = c(1, 2),
  labels = c("Sex code 1", "Sex code 2")
)

cat("Comb mass summary:\n")
summary(phenotype_data$comb_mass)

cat("\nBody weight at 212 days summary:\n")
summary(phenotype_data$body_weight)

cat("\nSex-group counts:\n")
table(phenotype_data$sex_group, useNA = "ifany")
```

```{r}
cat("Individuals with recorded comb mass:",
    sum(!is.na(phenotype_data$comb_mass)), "\n")

cat("Individuals with recorded body weight:",
    sum(!is.na(phenotype_data$body_weight)), "\n")

cat("Individuals with recorded sex:",
    sum(!is.na(phenotype_data$sex_code)), "\n")
```

#### Observations

Comb mass is available for 447 individuals, with 125 missing values. It ranges from 0.20 to 63.30 g, with a median of 6.00 g and a mean of 15.22 g. The middle 50% of measurements lie between 2.00 and 27.95 g.

Body weight at 212 days is available for 471 individuals, with 101 missing values. It ranges from 684 to 1,999 recorded units, with a median of 1,178 and a mean of 1,199.6. There are 272 individuals with sex code 1, 299 with sex code 2, and one with missing sex information.

#### Interpretation

The substantially higher mean than median for comb mass suggests an asymmetric distribution toward larger values, which should be examined using the histogram and sex-group plots. Body weight has a mean closer to its median, although this alone does not establish a symmetric or normal distribution.

### Examine the comb-mass distribution

```{r}
#| label: fig-comb-mass-distribution
#| fig-cap: "Distribution of the 447 recorded comb-mass measurements."

hist(
  phenotype_data$comb_mass,
  breaks = 25,
  col = "#0072B2",
  border = "white",
  main = "Distribution of chicken comb mass",
  xlab = "Comb mass (g)",
  ylab = "Number of individuals",
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)
```

#### **Observations**

The histogram shows a large concentration of comb masses below 6 g, with the tallest bar covering approximately 0–2 g. Frequencies decrease sharply above this range, followed by a broader concentration around 18–40 g. Relatively few individuals have comb masses above 45 g, and the maximum is 63.30 g. The distribution is asymmetric, consistent with the mean (15.22 g) exceeding the median (6.00 g).

#### **Interpretation**

The low-mass concentration and broader concentration at larger values suggest that the pooled distribution may combine groups with different comb-mass distributions. Comparing the recorded sex groups will help assess whether sex contributes to this pattern.

### Compare comb mass between sex groups

Sex groups retain their original codes because their biological labels have not yet been confirmed.

```{r}
#| label: fig-comb-mass-by-sex
#| fig-cap: "Comb mass grouped by the recorded sex codes."

boxplot(
  comb_mass ~ sex_group,
  data = phenotype_data,
  col = c("#0072B2", "#D55E00"),
  main = "Comb mass by sex group",
  xlab = "Recorded sex group",
  ylab = "Comb mass (g)",
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)
```

```{r}
#| label: comb-mass-summary-by-sex

sex_summary <- do.call(
  rbind,
  lapply(levels(phenotype_data$sex_group), function(group) {
    values <- phenotype_data$comb_mass[
      which(phenotype_data$sex_group == group)
    ]

    values <- values[!is.na(values)]

    data.frame(
      sex_group = group,
      observed = length(values),
      mean = mean(values),
      median = median(values),
      sd = sd(values)
    )
  })
)

sex_summary[, c("mean", "median", "sd")] <- round(
  sex_summary[, c("mean", "median", "sd")],
  2
)

print(sex_summary, row.names = FALSE)
```

#### **Observations**

Comb mass differs markedly between the recorded sex groups. Among 216 individuals with sex code 1, the mean is 28.92 g, the median is 28.50 g, and the standard deviation is 10.08 g. Among 231 individuals with sex code 2, the corresponding values are 2.42 g, 2.00 g, and 1.63 g. The difference in mean comb mass is 26.50 g. The boxplot shows substantially larger comb masses and greater absolute variability in sex code 1, with some overlap between the groups. Both groups contain observations above their upper whiskers.

#### **Interpretation**

The pronounced separation between sex groups helps explain the two concentrations in the pooled comb-mass histogram. Sex is therefore an important covariate to consider when comparing QTL scans with and without adjustment. Points beyond the boxplot whiskers are unusual within their group but do not, by themselves, indicate measurement errors. The groups remain labeled by their recorded codes until the male/female coding is confirmed.

### Examine the body-weight distribution

```{r}
#| label: fig-comb-body-weight-distribution
#| fig-cap: "Distribution of the 471 recorded body-weight measurements."

hist(
  phenotype_data$body_weight,
  breaks = 25,
  col = "#009E73",
  border = "white",
  main = "Body weight at 212 days",
  xlab = "Body weight (recorded units)",
  ylab = "Number of individuals",
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)
```

#### **Observations** 

Body weight at 212 days shows a broad distribution, with the tallest bars around 1,050–1,100 and 1,200–1,250 recorded units. Frequencies generally decrease toward both extremes, with relatively few individuals below 750 or above 1,750. The distribution has a modest right tail, consistent with the mean (1,199.6) being slightly higher than the median (1,178). The observed range is 684–1,999, based on 471 individuals.

#### **Interpretation**

Body weight shows less pronounced separation into distinct concentrations than pooled comb mass. The histogram describes variation in body weight but does not establish its relationship with comb mass. Examining their association overall and within each sex group will help assess the relevance of body weight as a covariate in the QTL scans.

### Examine comb mass in relation to body weight

The scatterplot uses individuals with complete comb-mass, sex, and body-weight measurements.

```{r}
#| label: prepare-comb-scatter-data

complete_phenotypes <- complete.cases(
  phenotype_data[, c("comb_mass", "body_weight", "sex_group")]
)

phenotype_complete <- phenotype_data[
  complete_phenotypes,
]

cat("Individuals with complete measurements:\n")
nrow(phenotype_complete)

sex_colors <- c("#0072B2", "#D55E00")
sex_symbols <- c(16, 17)
```

```{r}
#| label: fig-comb-mass-vs-body-weight
#| fig-cap: "Comb mass versus body weight for the 440 individuals with complete measurements, colored by sex code."
#| fig-width: 9
#| fig-height: 7

plot(
  phenotype_complete$body_weight,
  phenotype_complete$comb_mass,
  col = adjustcolor(
    sex_colors[as.integer(phenotype_complete$sex_group)],
    alpha.f = 0.6
  ),
  pch = sex_symbols[as.integer(phenotype_complete$sex_group)],
  main = "Comb mass in relation to body weight",
  xlab = "Body weight at 212 days (recorded units)",
  ylab = "Comb mass (g)",
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)

legend(
  "topleft",
  legend = levels(phenotype_complete$sex_group),
  col = sex_colors,
  pch = sex_symbols,
  bty = "n"
)
```

Calculate correlations both across the complete sample and within each sex group.

```{r}
#| label: comb-mass-body-weight-correlations

correlation_summary <- data.frame(
  group = c("All complete individuals", "Sex code 1", "Sex code 2"),
  individuals = c(
    nrow(phenotype_complete),
    sum(phenotype_complete$sex_group == "Sex code 1"),
    sum(phenotype_complete$sex_group == "Sex code 2")
  ),
  pearson_r = c(
    cor(
      phenotype_complete$comb_mass,
      phenotype_complete$body_weight
    ),
    vapply(levels(phenotype_complete$sex_group), function(group) {
      group_data <- phenotype_complete[
        phenotype_complete$sex_group == group,
      ]

      cor(group_data$comb_mass, group_data$body_weight)
    }, numeric(1))
  )
)

correlation_summary$pearson_r <- round(
  correlation_summary$pearson_r,
  3
)

print(correlation_summary, row.names = FALSE)
```

#### **Observations**

Among the 440 individuals with complete comb-mass, body-weight, and sex records, the overall Pearson correlation between comb mass and body weight is r = 0.740. Within each sex group, the correlations are substantially weaker: r = 0.264 for sex code 1 (215 individuals) and r = 0.256 for sex code 2 (225 individuals). The scatterplot shows that individuals with sex code 1 generally have both greater body weight and larger comb mass than those with sex code 2, although their body-weight ranges overlap.

#### **Interpretation**

The strong pooled correlation partly reflects differences between the sex groups. Within each group, body weight has a modest positive association with comb mass. These patterns support comparing QTL scans adjusted for sex, body weight, and both covariates: adjustment will assess QTL associations after accounting for these measured differences. The correlations alone do not establish that greater body weight causes larger comb mass.

## 5. Check marker order and estimate the genetic map

Per the practical the marker order is checked before estimating the genetic map. Pairwise recombination fractions are examined first, followed by a comparison of alternative local marker orders.

### List markers in their supplied order

```{r}
#| label: supplied-marker-order

marker_order <- do.call(
  rbind,
  lapply(names(comb_cross$geno), function(chr) {
    supplied_map <- comb_cross$geno[[chr]]$map

    data.frame(
      chromosome = chr,
      chromosome_index = seq_along(supplied_map),
      marker = names(supplied_map),
      supplied_position = as.numeric(supplied_map)
    )
  })
)

marker_order$genome_index <- seq_len(nrow(marker_order))

marker_order <- marker_order[
  ,
  c(
    "genome_index",
    "chromosome",
    "chromosome_index",
    "marker",
    "supplied_position"
  )
]

rownames(marker_order) <- NULL

print(marker_order, row.names = FALSE)
```

The supplied positions are retained as input information.

### Calculate pairwise recombination fractions

```{r}
#| label: estimate-comb-recombination

comb_cross_rf <- est.rf(comb_cross)
```

### Examine recombination across both chromosomes

```{r}
#| label: fig-comb-pairwise-recombination
#| fig-cap: "Estimated pairwise recombination fractions for markers on chromosomes 1 and 3, in the supplied order."
#| fig-width: 9
#| fig-height: 9

par(
  pty = "s",
  mar = c(5, 5, 4, 2) + 0.1,
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)

plotRF(
  comb_cross_rf,
  chr = c(1, 3),
  what = "rf",
  col.scheme = "viridis",
  main = "Pairwise recombination: chromosomes 1 and 3"
)
```

#### **Observations**

Markers 1–9 belong to chromosome 1, and markers 10–40 belong to chromosome 3. The blocks comparing markers between chromosomes are predominantly purple, indicating larger estimated recombination fractions and little apparent linkage between the two chromosomes.

Within chromosome 1, a relatively continuous yellow band follows the diagonal, with colors generally becoming darker farther from it. Chromosome 3 also shows groups of nearby markers with small recombination fractions, but its diagonal pattern is interrupted. In particular, genome-wide marker indices 12 (Gg_rs15777012) and 17 (3_23988469) show dark rows and columns contrasting with the surrounding regions of stronger linkage.

#### **Interpretation**

The overall pattern supports the supplied chromosome grouping, but the interruptions on chromosome 3 warrant closer examination of marker order and genotype quality. Marker 3_23988469 also has the highest missingness (45.28%), whereas Gg_rs15777012 has only 0.87% missingness. Thus, high missingness alone cannot explain all the unusual patterns. The chromosome-specific plots and alternative-order comparisons are needed before deciding whether to revise the marker order.

### Examine each chromosome separately

```{r}
#| label: fig-comb-recombination-chr1
#| fig-cap: "Estimated pairwise recombination fractions among the nine chromosome 1 markers."
#| fig-width: 9
#| fig-height: 9

par(
  pty = "s",
  mar = c(5, 5, 4, 2) + 0.1,
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)

plotRF(
  comb_cross_rf,
  chr = 1,
  what = "rf",
  col.scheme = "viridis",
  main = "Pairwise recombination: chromosome 1"
)
```

```{r}
#| label: fig-comb-recombination-chr3
#| fig-cap: "Estimated pairwise recombination fractions among the 31 chromosome 3 markers."
#| fig-width: 9
#| fig-height: 9

par(
  pty = "s",
  mar = c(5, 5, 4, 2) + 0.1,
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)

plotRF(
  comb_cross_rf,
  chr = 3,
  what = "rf",
  col.scheme = "viridis",
  main = "Pairwise recombination: chromosome 3"
)
```

#### **Observations** 

**Chromosome 1.** Recombination fractions generally increase as markers become farther apart in the supplied order: colors change from yellow or green near the diagonal toward purple farther away. Markers 5 (Gg_rs13939684) and 6 (Gg_rs15427022) form a particularly bright off-diagonal pair, indicating a small estimated recombination fraction. No similarly conspicuous dark row or column interrupts the neighboring-marker pattern.

**Chromosome 3.** Two markers clearly interrupt the diagonal pattern: 3 (Gg_rs15777012) and 8 (3_23988469). Marker 3 has dark-purple cells with neighboring markers 2 (Gg_rs15015376) and 4 (chr3_16000000). Marker 8 likewise shows darker cells with neighboring markers 7 (Gg_rs15282380) and 9 (Gg_rs14327472) than the surrounding diagonal blocks. These patterns indicate relatively large estimated recombination fractions despite proximity in the supplied order. Several other regions contain bright blocks, particularly around marker indices 13–16, indicating small recombination fractions among those markers.

#### **Interpretation**

Chromosome 1 appears broadly consistent with the supplied order, whereas chromosome 3 contains specific inconsistencies requiring further investigation. The patterns at Gg_rs15777012 and 3_23988469 could reflect incorrect placement or genotype-coding problems; the heatmap alone cannot distinguish these explanations. Although 3_23988469 has substantial missingness (45.28%), Gg_rs15777012 has little missingness (0.87%). Alternative-order comparisons and genotype checks are therefore needed before changing the map. Bright blocks also suggest that the data may provide limited information for resolving the order of some closely linked markers.

### Compare alternative local marker orders

The supplied order is compared with alternative orders within sliding windows of five markers. The comparison uses the minimum number of crossovers needed to explain the observed genotypes.

```{r}
#| label: check-comb-marker-order

ripple_chr1 <- ripple(
  comb_cross_rf,
  chr = "1",
  window = 5,
  method = "countxo"
)

ripple_chr3 <- ripple(
  comb_cross_rf,
  chr = "3",
  window = 5,
  method = "countxo"
)
```

```{r}
#| label: summarize-comb-marker-orders

cat("Chromosome 1: comparison of local marker orders\n")
print(
  unclass(summary(ripple_chr1)),
  max = 100000
)

cat("\nChromosome 3: comparison of local marker orders\n")
print(
  unclass(summary(ripple_chr3)),
  max = 100000
)
```

#### **Observations** 

**Chromosome 1.** The supplied order requires 1,261 obligate crossovers, compared with 1,297 after exchanging markers 5 and 6. No improvement was found in the five-marker-window search, supporting retention of the supplied order.

**Chromosome 3.** Of the 88 alternatives reported in the summary, 84 require fewer crossovers than the supplied order, three have the same count, and one requires two more. The supplied order requires 3,705 crossovers, whereas the best alternative requires 3,311, a reduction of 394 (10.6%). Its order begins 3–1–2–4–5, moving Gg_rs15777012 (marker 3) to the beginning of the chromosome. The second-best order begins 3–1–2–5–4 and requires 3,315 crossovers. The next three alternatives each require 3,373. All reported alternatives retain 3_23988469 (marker 8) in position 8.

#### **Interpretation**

The strongest improvement involves repositioning Gg_rs15777012, consistent with its unusual pattern in the recombination heatmap. The four-crossover difference between the two leading orders provides limited discrimination between the arrangements of markers 4 and 5. However, this local search does not establish the globally optimal chromosome order. Retention of 3_23988469 in every reported alternative does not confirm its placement: its unusual recombination pattern remains unresolved and requires further investigation. Allele coding should be checked before adopting a revised order.

### Check for possible reversed allele coding

This diagnostic checks whether reversing allele coding at a marker would increase its strongest pairwise linkage LOD score.

```{r}
#| label: check-comb-allele-coding

allele_check <- checkAlleles(
  comb_cross_rf,
  threshold = 3,
  verbose = TRUE
)

if (!is.null(allele_check) && nrow(allele_check) > 0) {
  print(allele_check, row.names = FALSE)
}
```

### Inspect recombination fractions around suspicious markers

The numerical estimates are examined for chromosome 3 markers Gg_rs15777012 and 3_23988469 and their immediate neighbors in the supplied order.

```{r}
#| label: inspect-suspicious-marker-recombination

rf_chr3 <- pull.rf(
  comb_cross_rf,
  chr = "3",
  what = "rf"
)

neighbor_pairs <- data.frame(
  focal_marker = c(
    "Gg_rs15777012",
    "Gg_rs15777012",
    "3_23988469",
    "3_23988469"
  ),
  neighboring_marker = c(
    "Gg_rs15015376",
    "chr3_16000000",
    "Gg_rs15282380",
    "Gg_rs14327472"
  )
)

neighbor_pairs$estimated_rf <- vapply(
  seq_len(nrow(neighbor_pairs)),
  function(i) {
    rf_chr3[
      neighbor_pairs$focal_marker[i],
      neighbor_pairs$neighboring_marker[i]
    ]
  },
  numeric(1)
)

neighbor_pairs$estimated_rf <- round(
  neighbor_pairs$estimated_rf,
  4
)

print(neighbor_pairs, row.names = FALSE)
```

#### **Observations**

The allele-coding check returned “No apparent problems,” meaning that no marker met the specified threshold of a three-unit increase in maximum pairwise linkage LOD after reversing its allele coding.

For Gg_rs15777012, estimated recombination fractions with its supplied neighbors are 0.5454 with Gg_rs15015376 and 0.5292 with chr3_16000000. For 3_23988469, the corresponding estimates are 0.6889 with Gg_rs15282380 and 0.0355 with Gg_rs14327472.

#### **Interpretation**

Gg_rs15777012 shows little evidence of linkage to its immediate neighbors under the supplied coding, supporting further investigation of its placement. In contrast, 3_23988469 has a strongly asymmetric neighboring-marker pattern: a high estimated fraction with Gg_rs15282380 but a small fraction with Gg_rs14327472. It is therefore not weakly linked to both neighbors.

Estimates above 0.5 should not be interpreted as genetic distances or as more than 50% observable recombination in a conventional linkage map. They warrant investigation of coding, placement, and the cross-model approximation. The negative allele check provides no support for automatically reversing the coding, but it does not rule out all genotype problems, particularly given the practical’s warning about linkage LOD behavior in this advanced intercross.

### Apply the candidate chromosome 3 order

The leading alternative from the initial local-order search is tested in a separate cross object.

```{r}
#| label: apply-candidate-chr3-order

candidate_order_chr3 <- c(
  3L,
  1L,
  2L,
  seq.int(4L, nmar(comb_cross_rf)[["3"]])
)

comb_cross_candidate <- switch.order(
  comb_cross_rf,
  chr = "3",
  order = candidate_order_chr3,
  error.prob = 0.01,
  map.function = "haldane"
)

candidate_marker_order <- data.frame(
  candidate_index = seq_along(candidate_order_chr3),
  original_index = candidate_order_chr3,
  marker = colnames(comb_cross_candidate$geno[["3"]]$data)
)

print(candidate_marker_order, row.names = FALSE)
```

### Compare crossover counts before and after rearrangement

```{r}
#| label: compare-candidate-crossover-counts

crossover_comparison <- data.frame(
  chromosome = c("1", "3"),
  supplied_order = c(
    sum(countXO(comb_cross_rf, chr = "1")),
    sum(countXO(comb_cross_rf, chr = "3"))
  ),
  candidate_order = c(
    sum(countXO(comb_cross_candidate, chr = "1")),
    sum(countXO(comb_cross_candidate, chr = "3"))
  )
)

crossover_comparison$reduction <- with(
  crossover_comparison,
  supplied_order - candidate_order
)

print(crossover_comparison, row.names = FALSE)
```

### Examine recombination under the candidate order

```{r}
#| label: fig-comb-candidate-recombination-chr3
#| fig-cap: "Pairwise recombination fractions on chromosome 3 after moving Gg_rs15777012 to the beginning of the marker order."
#| fig-width: 9
#| fig-height: 9

par(
  pty = "s",
  mar = c(5, 5, 4, 2) + 0.1,
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)

plotRF(
  comb_cross_candidate,
  chr = "3",
  what = "rf",
  col.scheme = "viridis",
  main = "Chromosome 3: candidate marker order"
)
```

### Check for further local-order improvements

The local-order search is repeated around the candidate order. This allows additional rearrangements to be assessed after the first change.

```{r}
#| label: recheck-candidate-chr3-order

ripple_chr3_candidate <- ripple(
  comb_cross_candidate,
  chr = "3",
  window = 5,
  method = "countxo"
)

print(
  unclass(summary(ripple_chr3_candidate)),
  max = 100000
)
```

#### **Observations**

Moving Gg_rs15777012 from position 3 to position 1 reduces the chromosome 3 crossover count from 3,705 to 3,311, a reduction of 394 (10.6%). Chromosome 1 remains unchanged at 1,261.

The repeated local-order search evaluates 2,616 orders and finds no further improvement. One alternative also requires 3,311 crossovers: it exchanges candidate positions 30 (Gg_rs16289999) and 31 (Gg_rs14368844). Another reported alternative exchanges positions 13 (Gg_rs15321001) and 14 (3_36969573) and requires 3,313 crossovers, two more than the candidate order.

The heatmap shows a more continuous region of small recombination fractions around positions 2–7. However, Gg_rs15777012, now at position 1, retains its dark row and column, and 3_23988469, still at position 8, continues to interrupt the neighboring-marker pattern.

#### **Interpretation**

The candidate order is a local minimum under the crossover-count criterion for the five-marker windows examined. The tied order indicates that this criterion cannot distinguish the order of the final two markers, so their current arrangement can be retained provisionally. Reordering improves the crossover count but does not resolve the unusual pairwise relationships of Gg_rs15777012 or 3_23988469. This order can therefore serve as a provisional working order for map estimation, with these marker uncertainties documented.

### Estimate genetic distances

Genetic distances are estimated using all available individuals and the candidate marker order. The assumed genotyping error probability is 0.01, with the Haldane map function.

```{r}
#| label: estimate-comb-genetic-map

comb_map_est <- est.map(
  comb_cross_candidate,
  error.prob = 0.01,
  map.function = "haldane",
  offset = 0,
  verbose = TRUE
)
```

### Summarize marker positions and chromosome lengths

```{r}
#| label: summarize-comb-genetic-map

estimated_marker_positions <- do.call(
  rbind,
  lapply(names(comb_map_est), function(chr) {
    positions <- comb_map_est[[chr]]

    data.frame(
      chromosome = chr,
      marker_index = seq_along(positions),
      marker = names(positions),
      position_cm = round(as.numeric(positions), 3)
    )
  })
)

rownames(estimated_marker_positions) <- NULL

print(estimated_marker_positions, row.names = FALSE)

map_length_summary <- do.call(
  rbind,
  lapply(names(comb_map_est), function(chr) {
    positions <- as.numeric(comb_map_est[[chr]])

    data.frame(
      chromosome = chr,
      markers = length(positions),
      length_cm = round(diff(range(positions)), 3)
    )
  })
)

cat("\nEstimated chromosome lengths:\n")
print(map_length_summary, row.names = FALSE)
```

### Inspect intervals between adjacent markers

```{r}
#| label: inspect-comb-map-intervals

map_intervals <- do.call(
  rbind,
  lapply(names(comb_map_est), function(chr) {
    positions <- comb_map_est[[chr]]
    marker_names <- names(positions)

    data.frame(
      chromosome = chr,
      left_marker = head(marker_names, -1),
      right_marker = tail(marker_names, -1),
      interval_cm = round(diff(as.numeric(positions)), 3)
    )
  })
)

map_intervals <- map_intervals[
  order(-map_intervals$interval_cm),
]

rownames(map_intervals) <- NULL

print(map_intervals, row.names = FALSE)
```

### Plot the estimated genetic map

```{r}
#| label: fig-comb-estimated-map
#| fig-cap: "Provisional genetic map for chromosomes 1 and 3, estimated under the intercross approximation using the candidate marker order."
#| fig-width: 8
#| fig-height: 9

par(
  pty = "m",
  mar = c(5, 5, 4, 2) + 0.1,
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)

plotMap(
  comb_map_est,
  shift = FALSE,
  show.marker.names = FALSE,
  main = "Estimated genetic map: chromosomes 1 and 3"
)
```

### Create a cross containing the estimated map

```{r}
#| label: replace-comb-genetic-map

comb_cross_mapped <- replace.map(
  comb_cross_candidate,
  comb_map_est
)
```

#### Observations and interpretation

The estimated map spans 151.72 cM on chromosome 1 and 1,390.55 cM on chromosome 3. Chromosome 3’s unusually large length is dominated by the 1,001.51 cM interval between Gg_rs15777012 and Gg_rs14084016, which accounts for approximately 72% of its total length. This explains the large empty region in the map plot.

Moving Gg_rs15777012 to the beginning of chromosome 3 reduced the number of inferred crossovers, but the resulting map shows that its placement remains problematic. A locally preferred marker order does not necessarily produce a reliable genetic map. Numerical convergence therefore does not establish that this interval represents a meaningful genetic distance.

Another large interval occurs between Gg_rs15282380 and 3_23988469 (101.51 cM). In contrast, 3_23988469 and Gg_rs14327472 are separated by only 1.99 cM. This asymmetry is consistent with the earlier pairwise recombination results. The substantial missingness at 3_23988469 should also be considered when interpreting its placement.

Two marker pairs have identical estimated positions at the displayed precision: Gg_rs16251229/Gg_rs15321001 and Gg_rs16289999/Gg_rs14368844. These positions indicate that the analysis did not resolve their separation; they do not, by themselves, establish that the markers have identical genotypes.

### Examine linkage of the problematic chromosome 3 marker

```{r}
#| label: inspect-problematic-marker

problem_marker <- "Gg_rs15777012"

# Use the preserved cross with the original marker order.
chr3_genotypes <- comb_cross_rf$geno[["3"]]$data
chr3_markers <- colnames(chr3_genotypes)

stopifnot(problem_marker %in% chr3_markers)

chr3_rf <- pull.rf(
  comb_cross_rf,
  what = "rf",
  chr = "3"
)

chr3_linkage_lod <- pull.rf(
  comb_cross_rf,
  what = "lod",
  chr = "3"
)

# Count individuals with nonmissing genotype calls at both markers.
problem_genotypes <- chr3_genotypes[, problem_marker]

jointly_typed <- vapply(
  chr3_markers,
  function(marker) {
    sum(
      !is.na(problem_genotypes) &
        !is.na(chr3_genotypes[, marker])
    )
  },
  integer(1)
)

candidate_chr3_markers <- colnames(
  comb_cross_candidate$geno[["3"]]$data
)

problem_marker_linkage <- data.frame(
  marker = chr3_markers,
  original_index = seq_along(chr3_markers),
  candidate_index = match(
    chr3_markers,
    candidate_chr3_markers
  ),
  jointly_typed = unname(jointly_typed),
  recombination_fraction = as.numeric(
    chr3_rf[problem_marker, chr3_markers]
  ),
  linkage_lod = as.numeric(
    chr3_linkage_lod[problem_marker, chr3_markers]
  )
)

# Remove the marker's comparison with itself.
problem_marker_linkage <- problem_marker_linkage[
  problem_marker_linkage$marker != problem_marker,
]

# Show the smallest recombination fractions first.
problem_marker_linkage <- problem_marker_linkage[
  order(
    problem_marker_linkage$recombination_fraction,
    -problem_marker_linkage$linkage_lod,
    na.last = TRUE
  ),
]

rownames(problem_marker_linkage) <- NULL

cat(
  "Pairwise linkage of", problem_marker,
  "with the other chromosome 3 markers:\n\n"
)

print(
  problem_marker_linkage,
  row.names = FALSE,
  digits = 4
)
```

#### **Observations and interpretation**

Gg_rs15777012 has its smallest estimated recombination fraction with 3_23988469 (0.2398), whereas its relationship with its current neighboring marker, Gg_rs14084016, has an estimated recombination fraction of 0.5532. This supports questioning its placement at the beginning of chromosome 3 and helps explain the extremely large first interval in the estimated map.

However, the relationship with 3_23988469 does not establish a precise alternative position. It is based on 312 jointly typed individuals, and comparable close linkage is not evident with the surrounding markers. Overall, 26 of the 30 pairwise estimates exceed 0.5. The marker therefore remains difficult to place reliably under the current analysis. These results alone do not establish a genotyping error or justify reversing its allele coding.

### Compare all placements of Gg_rs15777012

```{r}
#| label: compare-problem-marker-placements

problem_marker <- "Gg_rs15777012"

original_chr3_markers <- colnames(
  comb_cross_rf$geno[["3"]]$data
)

candidate_chr3_markers <- colnames(
  comb_cross_candidate$geno[["3"]]$data
)

stopifnot(
  problem_marker %in% original_chr3_markers,
  setequal(original_chr3_markers, candidate_chr3_markers)
)

# Keep the other markers in their current relative order.
remaining_markers <- candidate_chr3_markers[
  candidate_chr3_markers != problem_marker
]

remaining_indices <- match(
  remaining_markers,
  original_chr3_markers
)

problem_index <- match(
  problem_marker,
  original_chr3_markers
)

n_chr3_markers <- length(original_chr3_markers)

# Give every trial the same starting map: 1 cM intervals.
# These are starting values, not the final estimated distances.
placement_start_map <- pull.map(comb_cross_rf)

placement_start_map[["3"]] <- setNames(
  seq_len(n_chr3_markers) - 1,
  original_chr3_markers
)

placement_base <- replace.map(
  comb_cross_rf,
  placement_start_map
)

# Fit each possible placement.
placement_trials <- lapply(
  seq_len(n_chr3_markers),
  function(position) {
    
    trial_order <- append(
      remaining_indices,
      problem_index,
      after = position - 1L
    )
    
    ordered_markers <- original_chr3_markers[trial_order]
    
    trial_cross <- switch.order(
      placement_base,
      chr = "3",
      order = trial_order,
      error.prob = 0.01,
      map.function = "haldane",
      maxit = 10000
    )
    
    # Obtain the fitted map and its log-likelihood attribute.
    trial_map <- est.map(
      trial_cross,
      chr = "3",
      error.prob = 0.01,
      map.function = "haldane",
      offset = 0,
      maxit = 10000,
      verbose = FALSE
    )
    
    chr3_map <- trial_map[["3"]]
    
    trial_summary <- data.frame(
      position = position,
      left_marker = if (position == 1L) {
        "(start)"
      } else {
        ordered_markers[position - 1L]
      },
      right_marker = if (position == n_chr3_markers) {
        "(end)"
      } else {
        ordered_markers[position + 1L]
      },
      crossovers = sum(countXO(trial_cross, chr = "3")),
      length_cm = diff(range(chr3_map)),
      largest_interval_cm = max(diff(chr3_map)),
      loglik = as.numeric(attr(chr3_map, "loglik"))
    )
    
    list(
      order = trial_order,
      map = trial_map,
      summary = trial_summary
    )
  }
)

placement_comparison <- do.call(
  rbind,
  lapply(placement_trials, function(trial) trial$summary)
)

# Express each fit's likelihood deficit in base-10 units.
best_loglik <- max(placement_comparison$loglik)

placement_comparison$lod_below_best <- (
  best_loglik - placement_comparison$loglik
) / log(10)

# Rank by likelihood, retaining the position identifier.
placement_ranked <- placement_comparison[
  order(
    placement_comparison$lod_below_best,
    placement_comparison$crossovers,
    placement_comparison$position
  ),
]

rownames(placement_ranked) <- NULL
```

```{r}
#| label: summarize-marker-placement-comparison

cat("Marker placement and obligate crossover counts:\n\n")

print(
  placement_ranked[
    , c("position", "left_marker", "right_marker", "crossovers")
  ],
  row.names = FALSE
)

cat("\nMap fit and interval lengths for the same placements:\n\n")

placement_fit_display <- placement_ranked[
  , c(
    "position",
    "length_cm",
    "largest_interval_cm",
    "loglik",
    "lod_below_best"
  )
]

placement_fit_display[, -1] <- round(
  placement_fit_display[, -1],
  3
)

print(
  placement_fit_display,
  row.names = FALSE
)
```

#### **Observations and interpretation**

All 31 placements of Gg_rs15777012 were compared while holding the remaining markers in a fixed relative order. Placement at the beginning of chromosome 3 produced the fewest obligate crossovers (3,311). However, the beginning and end placements had the same fitted log likelihood at the displayed precision, and both retained a 1,001.51 cM interval. Placement immediately before 3_23988469 did not resolve this interval either.

Thus, the preferred ordering criterion does not provide a useful estimate of this marker’s genetic position under the current model. Shorter maps obtained from some internal placements had substantially poorer likelihoods. These findings support treating Gg_rs15777012 as unresolved rather than assigning it a confident map position. They do not establish that its genotypes are erroneous.

### Construct a provisional working map

```{r}
#| label: construct-provisional-working-map

comb_cross_map_filtered <- drop.markers(
  comb_cross_candidate,
  markers = "Gg_rs15777012"
)

comb_map_filtered <- est.map(
  comb_cross_map_filtered,
  error.prob = 0.01,
  map.function = "haldane",
  offset = 0,
  verbose = TRUE
)

comb_cross_mapped_filtered <- replace.map(
  comb_cross_map_filtered,
  comb_map_filtered
)

cat("Provisional map after excluding Gg_rs15777012:\n\n")

summaryMap(comb_cross_mapped_filtered)
```

```{r}
#| label: fig-provisional-working-map

#| fig-cap: "Provisional genetic map with Gg_rs15777012 excluded because its placement remained unresolved."
#| fig-width: 8
#| fig-height: 9

par(
  mar = c(5, 5, 4, 2) + 0.1,
  cex.main = 1.3,
  cex.lab = 1.2,
  cex.axis = 1.1
)

plotMap(
  comb_cross_mapped_filtered,
  shift = FALSE,
  show.marker.names = FALSE,
  main = "Provisional genetic map: chromosomes 1 and 3"
)####
```

#### Observations and interpretation

Excluding the unresolved marker Gg_rs15777012 reduced the estimated chromosome 3 map length from approximately 1,390.5 to 389.0 cM, removing the extreme 1,001.5 cM terminal interval. Chromosome 1 remained unchanged at approximately 151.7 cM. The provisional map contains 39 markers across the two chromosomes.

Marker spacing remains uneven. Chromosome 3 retains a 101.5 cM interval between Gg_rs15282380 and 3_23988469, while several other markers have closely spaced positions. Consequently, mapping resolution will vary across the chromosomes, with greater uncertainty in sparsely covered intervals.

This map will be used for subsequent analyses, with Gg_rs15777012’s exclusion documented as a limitation. Removing its problematic placement does not establish that its genotypes were erroneous or validate every remaining interval. The estimated distances also remain conditional on the F2 model used to approximate this advanced intercross.

## 6. Examine possible genotyping errors

Error LOD scores measure how inconsistent an observed genotype is with the surrounding marker data under the assumed map and cross model. High scores identify calls for further inspection. Closely linked markers can also produce high scores from genuine recombination events, so flagged calls should not automatically be treated as errors.

### Calculate error LOD scores

```{r}
#| label: calculate-genotyping-error-lods

comb_cross_errorlod <- calc.errorlod(
  comb_cross_mapped_filtered,
  error.prob = 0.01,
  map.function = "haldane"
)
```

### Identify and summarize flagged genotype calls

Error LOD \> 4 as an initial screening cutoff, consistent with the package’s documentation.

```{r}
#| label: summarize-flagged-genotype-calls

error_lod_cutoff <- 4

error_candidates <- top.errorlod(
  comb_cross_errorlod,
  cutoff = error_lod_cutoff,
  msg = FALSE
)

# Handle the possibility that no calls exceed the cutoff.
if (is.null(error_candidates)) {
  error_candidates <- data.frame(
    chr = character(),
    id = integer(),
    marker = character(),
    errorlod = numeric()
  )
}

if (nrow(error_candidates) > 0L) {
  error_candidates <- error_candidates[
    order(-error_candidates$errorlod),
  ]
  
  rownames(error_candidates) <- NULL
}

cat(
  "Number of genotype calls with error LOD >",
  error_lod_cutoff, ":\n"
)
print(nrow(error_candidates))

cat("\nFlagged calls by chromosome:\n")

print(
  table(
    factor(
      as.character(error_candidates$chr),
      levels = names(comb_cross_errorlod$geno)
    )
  )
)

cat("\nUp to 20 calls with the highest error LOD scores:\n\n")

print(
  head(error_candidates, 20),
  row.names = FALSE,
  digits = 5
)
```

```{r}
#| label: summarize-error-candidates-by-marker

if (nrow(error_candidates) > 0L) {
  
  error_marker_summary <- do.call(
    rbind,
    lapply(
      split(error_candidates, error_candidates$marker),
      function(marker_calls) {
        data.frame(
          chromosome = as.character(marker_calls$chr[1]),
          marker = as.character(marker_calls$marker[1]),
          flagged_calls = nrow(marker_calls),
          maximum_error_lod = max(marker_calls$errorlod)
        )
      }
    )
  )
  
  error_marker_summary <- error_marker_summary[
    order(
      -error_marker_summary$flagged_calls,
      -error_marker_summary$maximum_error_lod
    ),
  ]
  
  rownames(error_marker_summary) <- NULL
  
  cat("Flagged genotype calls summarized by marker:\n\n")
  
  print(
    error_marker_summary,
    row.names = FALSE,
    digits = 5
  )
  
} else {
  cat("No genotype calls exceeded the screening cutoff.\n")
}
```

#### Observations and interpretation

Five genotype calls exceeded the screening cutoff of error LOD 4. All occurred on chromosome 3; no chromosome 1 calls exceeded this cutoff. Four flagged calls involved Gg_rs16251229 and Gg_rs15321001 in the same two individuals, 34512 and 34134. These markers previously had identical estimated positions at the displayed precision, so their flags may reflect a shared local inconsistency rather than four independent errors. A further call was flagged at chr3_16000000 in individual 34308.

These calls require inspection alongside neighboring genotypes. Error LOD scores alone cannot distinguish genotyping errors from genuine recombination, marker-order problems, or limitations of the assumed model. The absence of flags on chromosome 1 does not establish that every genotype there is correct.

### Inspect flagged genotype patterns

The individual IDs, 34512, 34134, and 34308, are matched to the cross before passing individuals to `plotGeno()`

```{r}
#| label: locate-flagged-individuals-and-markers

flagged_ids <- unique(as.character(error_candidates$id))

cross_ids <- as.character(comb_cross_errorlod$pheno$id)

flagged_rows <- match(flagged_ids, cross_ids)

stopifnot(!anyNA(flagged_rows))

flagged_individuals <- data.frame(
  individual_id = flagged_ids,
  cross_row = flagged_rows
)

cat("Flagged individuals and their cross row indices:\n\n")

print(flagged_individuals, row.names = FALSE)

# Identify the flagged markers in the current, filtered map.
chr3_map <- pull.map(
  comb_cross_errorlod,
  chr = "3"
)[["3"]]

flagged_markers <- unique(
  as.character(error_candidates$marker)
)

flagged_marker_positions <- data.frame(
  marker = flagged_markers,
  marker_index = match(flagged_markers, names(chr3_map)),
  position_cm = round(
    as.numeric(chr3_map[flagged_markers]),
    3
  )
)

flagged_marker_positions <- flagged_marker_positions[
  order(flagged_marker_positions$marker_index),
]

cat("\nFlagged marker indices and estimated positions:\n\n")

print(flagged_marker_positions, row.names = FALSE)
```

Plot chromosome 3 for these three individuals

```{r}
#| label: fig-flagged-chr3-genotypes
#| fig-cap: "Chromosome 3 genotype patterns for the three individuals with flagged calls. The screening cutoff is error LOD > 4."
#| fig-width: 12
#| fig-height: 6
#| out-width: "100%"

previous_par <- par(
  oma = c(0, 0, 3, 0)
)

plotGeno(
  comb_cross_errorlod,
  chr = "3",
  ind = flagged_rows,
  horizontal = TRUE,
  include.xo = TRUE,
  cutoff = error_lod_cutoff,
  min.sep = 2,
  cex = 1.1
)

mtext(
  "Chromosome 3: genotype calls requiring inspection",
  side = 3,
  outer = TRUE,
  line = 1,
  font = 2,
  cex = 1.2
)

par(previous_par)
```

#### Observations and interpretation

In individual 34512, Gg_rs16251229 has an H genotype between A calls at neighboring markers. This isolated change is suspicious because it requires additional recombination events to explain the local pattern. Both Gg_rs16251229 and Gg_rs15321001 are flagged, although the visible isolated change occurs at Gg_rs16251229.

In individual 34134, the same marker pair marks an A-to-H transition, with H calls continuing at subsequent markers. This pattern could represent a genuine recombination event. However, Gg_rs16251229 and Gg_rs15321001 have the same estimated position at the displayed precision (277.806 cM), making their differing calls highly inconsistent with the fitted map. The paired flags therefore do not establish which call, if either, is erroneous.

In individual 34308, chr3_16000000 has a B genotype between A calls at Gg_rs15749695 and 3_16300000. This isolated change is another candidate genotyping error, but genotype validation would be needed to confirm it.

Overall, the five flags identify local inconsistencies requiring attention rather than five confirmed errors. Their interpretation remains conditional on marker spacing, the provisional map, and the cross model.

## 7. Calculate genotype probabilities

### Calculate conditional genotype probabilities

```{r}
#| label: calculate-genotype-probabilities

comb_cross_genoprob <- calc.genoprob(
  comb_cross_errorlod,
  step = 1,
  off.end = 0,
  error.prob = 0.01,
  map.function = "haldane",
  stepwidth = "fixed"
)
```

Conditional genotype probabilities were calculated on a 1 cM grid using the provisional genetic map, the Haldane map function, and an assumed genotyping error probability of 0.01. The observed genotype calls, including the five calls flagged in Section 5, were retained. These probabilities remain conditional on the F2 approximation used for this advanced intercross. The 1 cM grid defines the spacing of scan evaluations; it does not imply 1 cM mapping resolution.

### Check the probability arrays

Each chromosome’s probability array has dimensions:

individuals × evaluated positions × genotype states

The F2 model has three genotype states. Evaluated positions include intermediate scan positions as well as marker locations.

```{r}
#| label: summarize-genotype-probabilities

genoprob_summary <- do.call(
  rbind,
  lapply(
    names(comb_cross_genoprob$geno),
    function(chr) {
      
      probability_array <- comb_cross_genoprob$geno[[chr]]$prob
      
      probability_sums <- apply(
        probability_array,
        c(1, 2),
        sum
      )
      
      data.frame(
        chromosome = chr,
        individuals = dim(probability_array)[1],
        evaluated_positions = dim(probability_array)[2],
        genotype_states = dim(probability_array)[3],
        maximum_sum_deviation = max(
          abs(probability_sums - 1)
        )
      )
    }
  )
)

rownames(genoprob_summary) <- NULL

cat("Genotype probability array dimensions and normalization:\n\n")

print(
  genoprob_summary,
  row.names = FALSE,
  digits = 4
)

cat("\nNumber of observed markers retained:\n")

print(nmar(comb_cross_genoprob))
```

#### Observations and interpretation

Genotype probabilities were calculated for all 572 individuals at 160 positions on chromosome 1 and 419 positions on chromosome 3. Each position has three possible genotype states under the F2 model. The probabilities sum to one within numerical precision, and all 39 markers in the provisional map were retained. These checks confirm the expected array structure and normalization; they do not validate the underlying map or cross-model assumptions.

### Simulate genotypes for multiple imputation

```{r}
#| label: simulate-genotypes

set.seed(20261007)

comb_cross_imputed <- sim.geno(
  comb_cross_genoprob,
  n.draws = 64,
  step = 1,
  off.end = 0,
  error.prob = 0.01,
  map.function = "haldane",
  stepwidth = "fixed"
)
```

## 8. Compare single-QTL scans with and without covariates

### Define a common analysis sample

Select individuals with observed comb mass, sex, and body weight.

```{r}
#| label: define-common-scan-sample

scan_phenotypes_all <- pull.pheno(
  comb_cross_genoprob,
  pheno.col = c("comb_g", "sex", "w212")
)

scan_complete <- complete.cases(scan_phenotypes_all)

comb_cross_scan <- subset(
  comb_cross_genoprob,
  ind = scan_complete
)

scan_phenotypes <- pull.pheno(
  comb_cross_scan,
  pheno.col = c("comb_g", "sex", "w212")
)

cat("Individuals retained for all four scans:\n")
print(nind(comb_cross_scan))

cat("\nIndividuals excluded because of missing phenotype or covariate data:\n")
print(sum(!scan_complete))

cat("\nSex-group counts in the common analysis sample:\n")
print(table(scan_phenotypes$sex))
```

Encode sex as a binary covariate, retaining the neutral group labels because the sex-code key has not been independently confirmed.

```{r}
#| label: prepare-scan-covariates

stopifnot(
  all(scan_phenotypes$sex %in% c(1, 2))
)

scan_covariates <- cbind(
  sex_group_2 = as.numeric(scan_phenotypes$sex == 2),
  body_weight = scan_phenotypes$w212
)

stopifnot(
  nrow(scan_covariates) == nind(comb_cross_scan),
  !anyNA(scan_covariates)
)

head(scan_covariates)
```

All four scans use the same individuals with complete comb-mass, sex, and body-weight data. Sex isencoded as a binary indicator for group 2, with group 1 as the reference. Body weight is included as a continuous covariate. Both covariates are included as main effects, without interactions with QTL genotype.

#### Observations

The common analysis sample contains 440 individuals with complete comb-mass, sex, and body-weight data: 215 in sex group 1 and 225 in sex group 2. The remaining 132 individuals were excluded because at least one required value was missing. Using this same sample for all four scans prevents differences in sample composition from affecting the covariate comparison.
